Stochastic Variance-Reduced Policy Gradient

### 2018-02-09 svrpg.tex
added half-cheetah data, note on REINFORCE/PGT in rllab and minor fixes

\textit{1) Non-concavity.} A common assumption, in this case, is to assume the objective function to be $L$-smooth.
However, in RL we can consider the following assumption which is sufficient for the $L$-smoothness of the objective (see Lemma~\ref{lemma:lsmooth}).
	\begin{restatable}[On policy derivatives]{assumption}{boundedscore}\label{ass:bounded_score}
		For each state-action pair $(s,a)$, value of $\vtheta$ and dimensions $i,j$ there are positive constants $G,F<\infty$ such that:
\[
		\left|\nabla_{\theta_i}\log\pi_{\vtheta}(a\vert s)\right| \leq \GRADLOG, \qquad
        \left|\frac{\partial^2}{\partial\theta_i\partial\theta_j}\log\pi_{\vtheta}(a \vert s)\right| \leq \HESSLOG.
\]
	\end{restatable}

@@

We can now state the convergence guarantees for SVRPG.
\begin{restatable}[Convergence of the SVRPG algorithm]{theorem}{convergence}\label{theo:convergence}
Assume the REINFORCE or the G(PO)MDP gradient estimator is used in SVRPG (see Equation~\eqref{E:svrpg.estimate.batch}).
Under Assumptions~\ref{ass:bounded_score}, \ref{ass:REINFORCE} and \ref{ass:M2}, the parameter vector $\vtheta_A$ returned by Algorithm~\ref{alg:svrpg} after $T=m\times S$ iterations has, for some positive constants $\psi,\zeta, \xi$ and for proper choice of the step size $\alpha$ and the epoch size $m$, the following property:
\begin{align*}
	&\EVV[]
	{\norm[2]{\nabla J(\vtheta_A)}^2} 
		\leq
		\frac{J(\vtheta^*)-J(\vtheta_0)}{\psi T} +
		\frac{\zeta}{N}
		+\frac{\xi}{B},
\end{align*}
where $\psi,\zeta,\xi$ depend only on $\GRADLOG,\HESSLOG,\VARRF,\VARIS,\alpha$ and $m$.
\end{restatable}
% Specific constants and meta-parameter constraints are provided in the proof of the theorem in Appendix \ref{app:proofs}.
Refer to Appendix~\ref{app:proofs} for a detailed proof involving the definition of the constants and the meta-parameter constraints.
%The theoretical settings may be too conservative.
By analysing the upper-bound in Theorem~\ref{theo:convergence} we observe that: I) the $\mathcal{O}(\nicefrac{1}{T})$ term is coherent with results on non-concave SVRG~\citep[\eg][]{reddi2016stochastic}; II) the $\mathcal{O}(\nicefrac{1}{N})$ term is due to the FG approximation and is analogous to the one in~\citep{harikandeh2015stopwasting}; III) the $\mathcal{O}(\nicefrac{1}{B})$ term is due to importance weighting.
To achieve true convergence, both the batch size $N$ and the mini-batch size $B$ should increase over time. 
In practice, it is enough to choose $N$ and $B$ large enough to make the second and the third term negligible, \ie to mitigate the variance introduced by FG approximation and importance sampling, respectively.
Once the last two terms can be neglected, the number of trajectories needed to achieve $\norm[2]{\gradJ{\vtheta}}^2\leq\epsilon$ is $O(\frac{B+\nicefrac{N}{m}}{\epsilon})$. In this sense, an advantage over batch gradient ascent can be achieved with properly selected meta-parameters. In Section~\ref{sec:stopping} we propose a joint selection of step size $\alpha$ and epoch size $m$.
Finally, from the return statement of Algorithm \ref{alg:svrpg}, it is worth noting that $J(\vtheta_A)$ can be seen as the average performance of all the policies tried by the algorithm. This is particularly meaningful in the context of online learning that we are considering in this paper.

@@

\begin{figure}[t]
	\includegraphics[width=0.5\textwidth]{images/cart_pole_SVRPG_vs_GPOMDP}
	\caption{On-line performance over sampled trajectories of SVRPG vs G(PO)MDP in the Cart-pole task, with 90\% confidence intervals}
	\label{fig:cartpole}
\end{figure}

@@

\begin{table*}[t]
	\caption{
		\label{tab:1}
		Comparison between G(PO)MDP and SVRPG on different tasks. Average return over the whole learning process is reported, together with normalized standard deviation and 90\% confidence intervals.}
	\centering
	\begin{tabular}{@{}lccccccc@{}} 
		\toprule
		\phantom{abc} & \multicolumn{3}{c}{G(PO)MDP} & \phantom{abc} & \multicolumn{3}{c}{SVRPG} \\
		\cmidrule{2-4} \cmidrule{6-8}
		\phantom{abc} & Avg Return & Norm. Std & C.I. & \phantom{abc}
		& Avg Return & Norm. Std & C.I.
		\\\cmidrule{2-8}
		Cart-pole & 618.49 & 21.09 & [579.82, 657.16] & \phantom{abc} 
		& \textbf{736.31} & 16.51 & [706.04, 766.57] \\ 
		\phantom{abc} & \phantom{abc} & \phantom{abc} & \phantom{abc} & \phantom{abc} & \phantom{abc} & \phantom{abc} \\
		Swimmer & 39.86 & 3.84 & [32.82, \:\:\: 46.89] & \phantom{abc} & 49.60 & 3.27 & [43.60, \:\:\: 55.60]\\
		\bottomrule
	\end{tabular}
\end{table*}

@@

\paragraph{Preliminary results on actor-critic.}As mentioned, another variance-reduction technique in policy gradient is to employ baselines, or \textit{critics}. This tool is orthogonal to the methods described in this paper, and the theoretical results of Section \ref{sec:conv} are general in this sense. In the experiments described so far, we compared against the so-called \textit{actor-only} G(PO)MDP, \ie without baseline. In order to move towards a more general understanding of the problem of variance in policy gradient, we also test SVRPG in an \textit{actor-critic} scenario. To do so, we use the more challenging mujoco Half-cheetah task, a 3D locomotion task that has a larger state-action space than Swimmer. Table \ref{tab:1} also reports the comparison between SVRPG and G(PO)MDP on Half-cheetah using the critic suggested in \cite{duan2016benchmarking} for both algorithms. Results are promising, showing that a combination of baseline usage and SVRG-like variance reduction can yield an improvement that the two techniques alone are not able to achieve.

@@

\section{Policy Gradient Estimators} \label{A:gradient_estimators}
We present a brief overview of two most widespread gradient estimators (REINFORCE~\citep{williams1992simple} and G(PO)MDP~\citep{baxter2001infinite}) both in on-policy and off-policy settings.
Let $\tau = \{\langle s_t,a_t \rangle\}_{t=0}^{H}= \{z_t\}_{t=0}^{H} = z_{0:H}$ is a $(H+1)$-steps trajectory.
Given that $\tau$ depends on the MDP $M=\{\mathcal{S},\mathcal{A}, \mathcal{P}, \Reward, \gamma, \rho\}$ and the actual policy $\pi$, the trajectory is
said drawn from density distribution $p(\tau|\pi,M)$ defined as:
\[
        p(\tau|\pi,M) = \rho(s_0) \pi(z_0) \prod_{k=1}^{H} \mathcal{P}(s_k|z_{k-1})\pi(z_k).
\]
We can now recall the definition of policy performance 
\[
        J(\pol) = \EVV[\tau \sim p(\cdot|\pol)]{\Reward(\tau)|M},
\]
where $\Reward(\tau) = \sum_{t=0}^{H}\gamma^t \Reward(z_t)$.
The policy gradient $\nabla J(\vtheta)$ is
\begin{equation}\label{E:onpolicygradient}
        \gradJ{\vtheta} = \EVV[\tau \sim p(\cdot|\pol)]{\nabla \log\Delta(\tau|\vtheta)\Reward(\tau)} = \EVV[\tau \sim p(\cdot|\pol)]{\sum_{j=0}^{H} \gamma^j \mathcal{R}(z_j) \sum_{i=0}^{H} \nabla \log\pi(z_i)}.
\end{equation}

@@

\section{Experimental Details}\label{app:exp}
Used environments:
\begin{enumerate}
	\item \emph{Cart-Pole Balancing} : 4-dimensional state space: cart position x, pole angle $\theta$, cart velocity $\dot{x}$ and pole velocity $\dot{\theta}$; 1-dimensional action space: the horizontal force applied to the cart body. Reward function  is defined as $r(s, a) := 10 - (1 - cos(\theta)) - 10^{-5}\norm[] a^2$. The episodes terminates when $|x|>2.4$ or $|\theta|>0.2$ or the number of time steps T is greater than 100.
	\item \emph{Mujoco Swimmer}: 13-dimensional state space: 3 links velocities ($v_x$ and $v_y$ of center of masses) and 2 actuated joints angles. 2-dimensional action space: the two momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 10^{-4}\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
	\item \emph{Mujoco Half Cheetah}: 20-dimensional state space: 9 links and 6 actuated joints angles. 6-dimensional action space: the 6 momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 0.05\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
\end{enumerate}

Refer to \cite{duan2016benchmarking}  for more details about GPOMDP in Half Cheetah task.
All the experiments' parameters are reported in the following table:

\centering
\begin{tabular}{| l | c  c  c |}
	\hline	
	& Cart-Pole & Swimmer & Half Cheetah \\
	\hline
	State space dimension  & 4 & 13 & 20 \\
	Control space dimension & 1 & 2 & 6 \\
	NN Hidden layers & 8 & 32x32 & 100x50x25 \\
	Adam learning rate SVRPO & $5\cdotp10^{-2}$ & $10^{-3}$ & $10^{-3}$ \\
	Adam learning rate GPOMDP & $10^{-2}$ & $10^{-3}$ & $10^{-2}$ \\
	Snapshot batch size & 100 & 100 & 100 \\
	Sub-iterations batch size & 10 & 10 & 10 \\
	GPOMDP batch size & 10 & 10 & 100 \\
	Max number of sub-iterations & 50 & 20 & 20 \\
	Episodes horizon& 100 & 500 & 500 \\
	Baseline& No & No & Yes \\
	Discount $(\gamma)$& 0.99 & 0.995 & 0.99 \\
	Total number of trajectories& 10000 & 20000 & 50000 \\
	\hline  
\end{tabular}

>>>>

\textit{1) Non-concavity.} A common assumption, in this case, is to assume the objective function to be $L$-smooth.
However, in RL we can consider the following assumption which is sufficient for the $L$-smoothness of the objective (see Lemma~\ref{lemma:lsmooth}).
	\begin{restatable}[On policy derivatives]{assumption}{boundedscore}\label{ass:bounded_score}
		For each state-action pair $(s,a)$, any value of $\vtheta$ and all parameter components $i,j$ there are positive constants $G,F<\infty$ such that:
\[
		\left|\nabla_{\theta_i}\log\pi_{\vtheta}(a\vert s)\right| \leq \GRADLOG, \qquad
        \left|\frac{\partial^2}{\partial\theta_i\partial\theta_j}\log\pi_{\vtheta}(a \vert s)\right| \leq \HESSLOG.
\]
	\end{restatable}

@@

We can now state the convergence guarantees for SVRPG.
\begin{restatable}[Convergence of the SVRPG algorithm]{theorem}{convergence}\label{theo:convergence}
Assume the REINFORCE or the G(PO)MDP gradient estimator is used in SVRPG (see Equation~\eqref{E:svrpg.estimate.batch}).
Under Assumptions~\ref{ass:bounded_score}, \ref{ass:REINFORCE} and \ref{ass:M2}, the parameter vector $\vtheta_A$ returned by Algorithm~\ref{alg:svrpg} after $T=m\times S$ iterations has, for some positive constants $\psi,\zeta, \xi$ and for proper choice of the step size $\alpha$ and the epoch size $m$, the following property:
\begin{align*}
	&\EVV[]
	{\norm[2]{\nabla J(\vtheta_A)}^2} 
		\leq
		\frac{J(\vtheta^*)-J(\vtheta_0)}{\psi T} +
		\frac{\zeta}{N}
		+\frac{\xi}{B},
\end{align*}
where $\psi,\zeta,\xi$ depend only on $\GRADLOG,\HESSLOG,\VARRF,\VARIS,\alpha$ and $m$.
\end{restatable}
% Specific constants and meta-parameter constraints are provided in the proof of the theorem in Appendix \ref{app:proofs}.
Refer to Appendix~\ref{app:proofs} for a detailed proof involving the definition of the constants and the meta-parameter constraints.
%The theoretical settings may be too conservative.
By analysing the upper-bound in Theorem~\ref{theo:convergence} we observe that: I) the $\mathcal{O}(\nicefrac{1}{T})$ term is coherent with results on non-concave SVRG~\citep[\eg][]{reddi2016stochastic}; II) the $\mathcal{O}(\nicefrac{1}{N})$ term is due to the FG approximation and is analogous to the one in~\citep{harikandeh2015stopwasting}; III) the $\mathcal{O}(\nicefrac{1}{B})$ term is due to importance weighting.
To achieve asymptotic convergence, the batch size $N$ and the mini-batch size $B$ should increase over time. 
In practice, it is enough to choose $N$ and $B$ large enough to make the second and the third term negligible, \ie to mitigate the variance introduced by FG approximation and importance sampling, respectively.
Once the last two terms can be neglected, the number of trajectories needed to achieve $\norm[2]{\gradJ{\vtheta}}^2\leq\epsilon$ is $O(\frac{B+\nicefrac{N}{m}}{\epsilon})$. In this sense, an advantage over batch gradient ascent can be achieved with properly selected meta-parameters. In Section~\ref{sec:stopping} we propose a joint selection of step size $\alpha$ and epoch size $m$.
Finally, from the return statement of Algorithm \ref{alg:svrpg}, it is worth noting that $J(\vtheta_A)$ can be seen as the average performance of all the policies tried by the algorithm. This is particularly meaningful in the context of online learning that we are considering in this paper.

@@

\begin{figure}[t]
	\includegraphics[width=0.5\textwidth]{images/cart_pole_SVRPG_vs_GPOMDP}
	\caption{On-line performance over sampled trajectories of SVRPG vs G(PO)MDP in the Cart-pole task, with 90\% t-student confidence intervals. Results denoted with $\dagger$ are taken from \cite{duan2016benchmarking}.}
	\label{fig:cartpole}
\end{figure}

@@

\begin{table*}[t]
	\caption{
		\label{tab:1}
		Comparison between G(PO)MDP and SVRPG on different tasks. Average return over the whole learning process is reported, together with normalized standard deviation and 90\% confidence intervals. Results denoted with $\dagger$ are taken from \cite{duan2016benchmarking}.}
	\centering
	\begin{tabular}{@{}lccccccc@{}} 
		\toprule
		\phantom{abc} & \multicolumn{3}{c}{G(PO)MDP} & \phantom{abc} & \multicolumn{3}{c}{SVRPG} \\
		\cmidrule{2-4} \cmidrule{6-8}
		\phantom{abc} & Avg Return & Norm. Std & C.I. & \phantom{abc}
		& Avg Return & Norm. Std & C.I.
		\\\cmidrule{2-8}
		Cart-pole & $618.49$ & $21.09$ & $[579.82, 657.16]$ & \phantom{abc} 
		& $736.31$ & $16.51$ & $[706.04, 766.57]$ \\
		Swimmer & $39.86$ & $3.84$ & $[32.82, 46.89]$ & \phantom{abc} & $49.60$ & $3.27$ & $[43.60, 55.60]$\\
		Half-cheetah & $1183.1^\dagger$ & $69.20^\dagger$ & $[1035.58, 1330.62]^\dagger$ & \phantom{abc} & $1480.95$ & $73.21$ & $[1324.87, 1637.03]$\\
		\bottomrule
	\end{tabular}
\end{table*}

@@

\paragraph{Preliminary results on actor-critic.}As mentioned, another variance-reduction technique in policy gradient is to employ baselines, or \textit{critics}. This tool is orthogonal to the methods described in this paper, and the theoretical results of Section \ref{sec:conv} are general in this sense. In the experiments described so far, we compared against the so-called \textit{actor-only} G(PO)MDP, \ie without baseline. In order to move towards a more general understanding of the problem of variance in policy gradient, we also test SVRPG in an \textit{actor-critic} scenario. To do so, we use the more challenging mujoco Half-cheetah task, a 3D locomotion task that has a larger state-action space than Swimmer. Table \ref{tab:1} also reports the comparison between SVRPG and G(PO)MDP\footnote{\cite{duan2016benchmarking} report results on REINFORCE. However, inspection on \textit{rllab} code and documentation reveals that it is actually PGT \cite{sutton2000policy}, which is equivalent to G(PO)MDP \citep[shown by][]{peters2008reinforcement}. Using the name REINFORCE in a general way is inaccurate, but widespread.} on Half-cheetah using the critic suggested in \cite{duan2016benchmarking} for both algorithms. Results are promising, showing that a combination of baseline usage and SVRG-like variance reduction can yield an improvement that the two techniques alone are not able to achieve.

@@

\section{Policy Gradient Estimators} \label{A:gradient_estimators}
We present a brief overview of the two most widespread gradient estimators (REINFORCE~\citep{williams1992simple} and G(PO)MDP~\citep{baxter2001infinite}) both in on-policy and off-policy settings.
Let $\tau = \{\langle s_t,a_t \rangle\}_{t=0}^{H}= \{z_t\}_{t=0}^{H} = z_{0:H}$ is a $(H+1)$-steps trajectory.
Given that $\tau$ depends on the MDP $M=\{\mathcal{S},\mathcal{A}, \mathcal{P}, \Reward, \gamma, \rho\}$ and the actual policy $\pi$, the trajectory is
said drawn from density distribution $p(\tau|\pi,M)$ defined as:
\[
        p(\tau|\pi,M) = \rho(s_0) \pi(z_0) \prod_{k=1}^{H} \mathcal{P}(s_k|z_{k-1})\pi(z_k).
\]
We can now recall the definition of policy performance 
\[
        J(\pol) = \EVV[\tau \sim p(\cdot|\pol)]{\Reward(\tau)|M},
\]
where $\Reward(\tau) = \sum_{t=0}^{H}\gamma^t \Reward(z_t)$.
The policy gradient $\nabla J(\vtheta)$ is
\begin{equation}\label{E:onpolicygradient}
        \gradJ{\vtheta} = \EVV[\tau \sim p(\cdot|\pol)]{\nabla \log\Delta(\tau|\vtheta)\Reward(\tau)} = \EVV[\tau \sim p(\cdot|\pol)]{\sum_{j=0}^{H} \gamma^j \mathcal{R}(z_j) \sum_{i=0}^{H} \nabla \log\pi(z_i)}.
\end{equation}

@@

\section{Experimental Details}\label{app:exp}
Setting specifications for all the used environments:
\begin{enumerate}
	\item \emph{Cart-Pole Balancing} : 4-dimensional state space: cart position x, pole angle $\theta$, cart velocity $\dot{x}$ and pole velocity $\dot{\theta}$; 1-dimensional action space: the horizontal force applied to the cart body. Reward function  is defined as $r(s, a) := 10 - (1 - cos(\theta)) - 10^{-5}\norm[] a^2$. The episodes terminates when $|x|>2.4$ or $|\theta|>0.2$ or the number of time steps T is greater than 100.
	\item \emph{Mujoco Swimmer}: 13-dimensional state space: 3 links velocities ($v_x$ and $v_y$ of center of masses) and 2 actuated joints angles. 2-dimensional action space: the two momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 10^{-4}\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
	\item \emph{Mujoco Half Cheetah}: 20-dimensional state space: 9 links and 6 actuated joints angles. 6-dimensional action space: the 6 momentums applied on actuated joints.  The reward function is defined as $r(s, a) := v_x - 0.05\norm[2] a^2$. The episodes terminates when the number of time steps T is greater than 500.
\end{enumerate}

All the experiments' parameters are reported in the following table:

\centering
\begin{tabular}{| l | c  c  c |}
	\hline	
	& Cart-Pole & Swimmer & Half Cheetah \\
	\hline
	State space dimension  & 4 & 13 & 20 \\
	Control space dimension & 1 & 2 & 6 \\
	NN Hidden layers & 8 & 32x32 & 32x32 \\
	Learning rate SVRPO & $5\cdotp10^{-2}$ & $10^{-3}$ & $10^{-3}$ \\
	Learning rate GPOMDP & $10^{-2}$ & $10^{-3}$ & $10^{-3}$ \\
	Snapshot trajectories & 100 & 100 & 100 \\
	Sub-iterations trajectories & 10 & 10 & 10 \\
	Max sub-iterations & 50 & 20 & 20 \\
	Trajectories lengths& 500 & 500 & 500 \\
	Discount $(\gamma)$& 0.99 & 0.995 & 0.99 \\
	Total trajectories& 10000 & 20000 & 50000 \\
	\hline  
\end{tabular}
